### 2020-07-06 chapters/abstract.tex
Final changes and submission

%Approach/Solution and %Main contributions
In this thesis, we design a method that makes use of lightweight streaming aggregations and distribution divergence functions to alert about deviations in data patterns in real-time, while in the presence of large volumes of high velocity, highly skewed and seasonal data. We develop a two-phased and two-windowed method: first, we perform batch analysis on a reference window and then a stream analysis over a sliding streaming window. We aggregate the contents of both windows using an approximate histogram aggregation based on Exponential Moving Averages (EMAs). We do this for each event and each of its fields, also denominated features in our context. We measure the divergence between both reference and sliding window EMA-based histograms for each feature with the Jensen–Shannon Divergence function, but other distance functions could be used. For each feature, we find out the probability value of each divergence measure and apply the Holm-Bonferroni multiple test correction to find out which probability values are lowest and their associated features. We generate alerts based on a user-defined probability threshold, for each feature.

We evaluate our method through a series of tests, some with synthetic datasets and some with real data. We further split the tests into single and multi-feature analysis. Our method accurately detected the introduced anomalies in the synthetic dataset experiments while maintaining high throughput, for single and multi-feature analysis. However, experiments with real data were not as accurate. Despite not knowing the specific reasons for that, we defer further investigation to future work and formulate a set of hypotheses that might explain it and hence are worthy of pursuing next. Our set of experiments supports the claim that sliding window aggregations and distribution divergence methods can be combined to detect data pattern shifts in streaming scenarios with constant time and memory complexity.

>>>>

%Approach/Solution and %Main contributions
In this thesis, we design a method that makes use of lightweight streaming aggregations and distribution divergence functions to alert about deviations in data patterns in real-time, while in the presence of large volumes of high velocity, highly skewed and seasonal data. We develop a two-phased and two-windowed method: first, we perform a batch analysis on a reference window and then a stream analysis over a sliding streaming window. We aggregate the contents of both windows using an approximate histogram aggregation based on Exponential Moving Averages (EMAs). We do this for each event and each of its fields, also denominated features in our context. We measure the divergence between both reference and sliding window EMA-based histograms for each feature with the Jensen–Shannon Divergence function, but other distance functions could be used. For each feature, we find out the probability value of each divergence measure and apply the Holm-Bonferroni multiple test correction to find out which probability values are lowest and their associated features. We generate alerts based on a user-defined probability threshold, for each feature.

We evaluate our method through a series of tests, some with synthetic datasets and some with real data. We further split the tests into single and multi-feature analysis. Our method accurately detected the introduced anomalies in the experiments using synthetic datasets while maintaining high throughput, for single and multi-feature analysis. However, experiments with real data were not as accurate. Despite not knowing the specific reasons for that, we defer further investigation to future work and formulate a set of hypotheses that might explain it and hence are worthy of pursuing next. Our set of experiments supports the claim that sliding window aggregations and distribution divergence methods can be combined to detect data pattern shifts in streaming scenarios with constant time and memory complexity.

### 2020-07-06 chapters/background.tex
Final changes and submission

In this chapter, we discuss fundamental concepts that will help us identify the set of relevant techniques that are suitable to address the problem to be formulated and tackled in this thesis. In Section \ref{sec:stream-superset} we introduce batch and stream processing as two large scale data processing techniques. In Section \ref{sec:windows} we review windowing mechanisms that essentially turn unbounded data streams to finite datasets. In Section \ref{sec:aggregations} we introduce aggregations as a whole. Then, we specifically discuss sliding window aggregations, algorithms used for its computation and aggregation properties. In Section \ref{sec:emas} we present Exponential Moving Averages as aggregators that replace the sliding window contents for a lighter representation, providing an approximate estimation of the aggregation value.

@@

Batch processing is the processing of data grouped by batches. A batch is a collection of data points that have been grouped by certain criteria. Examples of such criteria could be \textit{"all the transactions from the past two months"}, \textit{"the last million transactions"} or even \textit{"all the transactions made by an user until now"}. Batch datasets are known as static data \cite{Martin-Batch-Defin}. An example application of such a technique would be processing all the keywords searched for in the google search engine, and then grouping the results by keyword to count how many times a keyword was used in searches (for trend analysis purposes). Between batch and stream processing, batch processing is the older paradigm. Historically, Apache Hadoop \cite{borthakur2007hadoop, Hadoop, ApacheHadoop} was one of the first widely known frameworks for large-scale batch processing, created in 2005. Hadoop uses a programming model called MapReduce \cite{MapReduce}. This paradigm applies a map function that processes key/value pairs and generates sets of intermediate ones, ultimately merging all intermediate values
using a reducer function. To do this, the distributed computation framework reads and writes from disk on intermediate steps, which makes it inefficient for applications that reuse working sets of data across multiple parallel operations (\textit{e.g.}, iterative Machine Learning algorithms). Nearly a decade later, Apache Spark \cite{ApacheSpark, Spark} was released as the industry accepted successor.
Spark retains the scalability and fault tolerance of MapReduce but introduces Resilient Distributed Datasets (RDDs) \cite{SparkRDDs}. An RDD is a read-only collection of objects partitioned across a set of machines that can be rebuilt if a partition is lost.
Additionally, Spark processes data in random access memory (RAM), while Hadoop MapReduce persists data to the disk after a map or reduce action. Thus, Spark achieves lower processing times.

@@

In order to provide a streaming solution, Apache Spark created the Spark Streaming \cite{SparkStreaming} module. Being a batch system in its core, Spark implements streaming as micro-batching \cite{SparkStreamingPaper}. This approach handles streaming data by grouping incoming events into very small batches. Then, to process the entire defined window of data, it joins all micro-batches into a larger one and performs the desired computations. However, this creates an artificial barrier that does not truly exist in the streaming context. More than conceptually different from a true stream processing system, processing a stream in a batch fashion, even if in small ones, presents some issues. First, since we are indeed working with batches, results will never be updated in real-time per event being processed. For example, if we use a batch size of 100 events, the system would only produce accurate results every 100 events. To handle system inactivity usually these micro-batch systems also have the notion of updating the aggregations with incomplete batches, if no event has entered the system after a pre-specified time period has elapsed. As such, micro-batching was created as a natural implementation of stream processing using the available batch processing systems of the time, but can not be considered true stream processing.

The Venn diagram in Figure \ref{fig:stream-superset} \cite{batch-is-a-special-case-of-streaming} illustrates the relation of both processing techniques. A data stream is unbounded or infinite while a bounded data stream has a beginning and an end. Using these definitions, the concept of an unbounded data stream contains that of a bounded dataset. That is, from an infinite stream, it is possible to fix a left and right limits to obtain a bounded and finite dataset. This relation is visible in Figure \ref{fig:unbounded-bounded-superset} from the same blog post.

@@

In Section \ref{sec:windows} we will cover exactly how an unbounded dataset (also known as a data stream) can be converted into a bounded dataset with the use of windows.

@@

In Section \ref{sec:stream-superset} we have defined a data stream as an unbounded dataset. But what exactly are the issues of working with an infinite stream of data? Why might we need to convert it into a bounded dataset and how to do so? In this section, we will analyze such issues and present sliding windows as the solution.

@@

Botan \emph{et al.} \cite{Botan-SECRET} define a window as:
\begin{definition}
A window W over a stream $\mathbb{S}$ is a finite subset of $\mathbb{S}$
\end{definition}
Windows can be categorized by two dimensions: the window size $W$ and the sliding step size $S$. Windows can be defined as tuple or time-based. Tuple-based windows are windows whose size is defined by \emph{number} of items contained in the window, whereas time-based windows are windows whose size is defined by the total \emph{time} a window can store. For instance, a tuple-based window of size $10^3$ tuples will store exactly $10^3$ items. In contrast, a time-based window storing the last 5 minutes of events can contain a varying number of tuples but will always keep the events arrived in the last 5 minutes. A window slides over incoming data and groups it, enabling us to compute aggregations over a defined and bounded set of events. The step size \textit{S} of a windowing technique is defined as the distance between consecutive windows \cite{Botan-SECRET}. This means that two consecutive windows W\textsubscript{1} and W\textsubscript{2} are exactly \textit{S} units away. As mentioned, these units can be units of time or a number of tuples.

@@

% tumbling: step = window size
A tumbling window is a window with a stepping size as large as the window, \textit{i.e.}, $S = W$. Hence, tumbling windows partition the stream into non-overlapping chunks with the same size. Since the intersection of different tumbling windows is always empty, each event is said to belong to exactly one tumbling window. An example of a tumbling window can be seen in Figure \ref{fig:tumbling-window} with a tuple-based window of size \textit{W} of three and an equally large step size \textit{S}.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=1.1]{figures/tumbling.pdf}
      \caption[Tumbling window]{Tumbling window, W=S=3}
      \label{fig:tumbling-window}
    \end{center}
\end{figure}

@@

An aggregation --- also known as a fold, reduce or accumulate function --- is an operation that reduces a collection of values into a single one. Examples of such are the computation of dataset sums, minima, averages or the number of distinct elements (cardinality). An aggregation may be exact or approximate. Exact aggregations compute a value from a sequence of elements, where no error or loss of resolution exists. On the other hand, approximate aggregations have an error margin to the associated aggregation value which is usually controllable by adjusting a set of parameters. 
Approximate aggregations are most useful in situations where memory is limited, the volume of data to process is large and/or computing the exact aggregation can not be done in useful time. Thus, approximate aggregators trade precision of results for efficient time and space usage. In Section \ref{sec:pds} we explore approximate aggregators that make such a trade-off.

Aggregations are present in most programming languages and frameworks. Quoting from the Java programming language documentation on the \textit{Stream.reduce()} method \cite{Java13StreamReduce}: 
\textit{"A reduction operation (also called a fold) takes a sequence of input elements and combines them into a single summary result by repeated application of a combining operation"}. Similarly, the MapReduce \cite{MapReduce} paradigm consists of first applying \textit{map} steps to the collection of data and then a \textit{reduce} one. \textit{Map} tasks will transform each item as desired and the \textit{reduce} task will merge the collection into a single value. Reduce methods make use of a \textit{combine} function and recursively apply it. The \textit{combine} function takes the current aggregation value and combines it with the next item in the set, returning the new aggregated state. Doing this recursively for each item in the collection results in a single value: the aggregate value. The combine function is from here onwards represented as $\oplus$ and the identity or neutral element of the aggregation as $\overline{\theta}$.

@@

While some aggregations can be computed on an infinite dataset, often we want to compute these over a finite one. As we have seen in Section \ref{sec:windows}, we can obtain a bounded dataset from a data stream by applying windowing techniques. When doing so, we obtain a finite dataset denoted as a sliding window and compute aggregations over it. Tangwongsan \emph{et al.} define such aggregations over sliding windows as Sliding Window Aggregations (SWAGs) \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}. As an example of a SWAG, consider the data center temperature monitoring problem from Section \ref{sec:windows}. In this example, we wanted to compute the average temperature per rack in the data center in the last 5 minutes. If we translate this SWAG into streaming SQL following a similar syntax to the one provided by Apache Flink \cite{ApacheFlink} we would get:

\begin{verbatim}
SELECT rack_id, AVG(temperature)
FROM Temperatures
GROUP BY TUMBLE(timestamp, Time.minutes(5), rack_id)
\end{verbatim}

In the query above, the
\texttt{TUMBLE(timestamp, Time.minutes(5), rack\_id)} statement groups incoming data using time-based tumbling windows of 5 minutes duration, using the \texttt{timestamp} field to evict events not in this time interval. The resulting list of rows has \texttt{rack\_id} as its key and as corresponding value a list of temperatures, from the past 5 minutes. Finally, for each rack and its list of temperatures, we compute an \texttt{AVG} aggregation, resulting in the average temperature per rack in the past 5 minutes. Note that this is very similar to any aggregation made in a standard Database Management System (DBMS). The main difference is that in standard SQL the \texttt{GROUP BY} clause includes only the columns on which the data will be grouped while in streaming SQL the \texttt{GROUP BY} also specifies the type of window used (\textit{e.g.}, in this case, a tumbling window).

@@

Two of the most basic SWAG algorithms are Recalculate-From-Scratch (RFS) and Subtract-On-Evict (SOE). Recalculate-From-Scratch (RFS) is the simplest and most general SWAG algorithm. As the name suggests, this algorithm recomputes the aggregation over the entire window for each update. For instance, the computation of a sum over a sliding window using RFS means that whenever the window changes all elements are summed. This approach works for every aggregation. However, given \textit{n} as the window size, this algorithm takes \textit{O(n)} space and time complexity which makes it unfit in most scenarios of large scale data processing. Recall that the combine function is represented by $\oplus$ and the identity or neutral element of the aggregation by $\overline{\theta}$. Pseudocode for the Recalculate-From-Scratch algorithm is shown in Algorithm \ref{pseudo:rfs}, assuming all functions have access to a \textit{vals} array which holds the window contents.

@@

Subtract-On-Evict (SOE) on the other hand, is a SWAG algorithm that incrementally computes the aggregation value. For each element inserted or evicted from the window, the aggregation value is updated without recomputing the entire aggregation like in RFS. SOE relies on expiring the evicted elements from the aggregation state. Hence, SOE only works for \textit{invertible} aggregations, that is, a function $\ominus$ exists such that $(x \oplus y) \ominus y = x$, for all $x$ and $y$. To illustrate the differences between SOE and RFS, consider the computation of a sum aggregation over a sliding window. The total sum of the window will be the aggregation state or value. When inserting an element, the updated total window sum equals the current total sum plus ($\oplus$) the new element. The sum aggregation is considered \textit{invertible} because, when evicting an element, the new aggregation state is the total sum computed thus far minus ($\ominus$) the evicted element. With window size of \textit{n}, SOE has \textit{O(n)} space complexity and \textit{O(1)} time complexity. However, aggregations are not always invertible thus making this algorithm very restrictive and not always applicable. Algorithm \ref{pseudo:soe} shows pseudocode for SOE assuming all functions have access to the window contents through the \textit{vals} array and access to the current aggregation state \textit{agg}.

@@

As it turns out, besides \textit{invertibility}, other properties allow us to group aggregations in different families. Tangwongsan \emph{et al.} construct Table \ref{tbl:aggregations-properties} to summarize the aggregation families and satisfying properties \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}.

To understand these properties, we must be familiar with the \textit{combine} and \textit{query} functions. The already mentioned \textit{combine} function merges the incoming event with the incrementally computed aggregation state. The \textit{query} function is responsible for returning a result for a given query according to the current aggregation state. For instance, consider once again the use case of computing a sum aggregation over a window of data. The \textit{combine} function would be the arithmetic operator \textit{+} and would merge the current aggregation state --- \textit{i.e.}, the current sum value --- with the new incoming event. The \textit{query} function would return the current aggregation state --- \textit{i.e.}, return the total sum computed.

Based on the \textit{combine} and \textit{query} functions, the following are formal definitions of these properties. An aggregation is said to have:

\begin{enumerate}
    \item  an \textbf{\textit{invertible combine}} if a function $\ominus$ exists such that $(x \oplus y) \ominus y = x$, for all $x$ and $y$

    \item  an \textbf{\textit{associative combine}} if $(x \oplus y) \oplus z = x \oplus (y \oplus z)$, for all $x$ and $y$

    \item  a \textbf{\textit{commutative combine}} if $x \oplus y = y \oplus x$, for all $x$ and $y$

    \item  a \textbf{\textit{size-preserving combine}} if the aggregation state resulting from $\oplus$ always occupies the same space in memory

    \item  a \textbf{\textit{unary query}} if the query does not need any parameters --- \textit{e.g.}, a membership test with a Bloom filter would require the element to test membership for as a parameter so it is not unary; returning the total sum is unary since it does not require any parameters
\end{enumerate}

@@

    \end{center}
    \caption[Aggregation properties]{Aggregation properties. Checkmarks (\checkmark), crosses ($\times$), and question marks (?) indicate a property is true for all, false for all, or false for some of a given group of like operations, respectively}
  \label{tbl:aggregations-properties}
\end{table}

@@

In Section \ref{sec:windows} we discussed windowing mechanisms that allowed us to convert unbounded data streams (or infinite datasets) to finite datasets.

In Section \ref{sec:aggregations} we introduced aggregations over datasets. We gave examples of computable aggregations over unbounded datasets at first, but then discussed sliding window aggregations, which make use of windowing techniques to obtain a finite dataset, on which the aggregation is computed. We further discussed algorithms used for the efficient computation of sliding window aggregations. Lastly, we presented a set of aggregation properties.

>>>>

In this chapter, we discuss fundamental concepts that will help us identify the set of relevant techniques that are suitable to address the problem to be formulated and tackled in this thesis. In Section \ref{sec:stream-superset}, we introduce batch and stream processing as two large scale data processing techniques. In Section \ref{sec:windows}, we review windowing mechanisms that essentially turn unbounded data streams to finite datasets. In Section \ref{sec:aggregations}, we introduce aggregations as a whole. Then, we specifically discuss sliding window aggregations, algorithms used for its computation and aggregation properties. In Section \ref{sec:emas}, we present Exponential Moving Averages as aggregators that replace the sliding window contents for a lighter representation, providing an approximate estimation of the aggregation value.

@@

Batch processing is the processing of data grouped by batches. A batch is a collection of data points that have been grouped by certain criteria. Examples of such criteria could be \textit{"all the transactions from the past two months"}, \textit{"the last million transactions"} or even \textit{"all the transactions made by an user until now"}. Batch datasets are known as static data \cite{Martin-Batch-Defin}. An example application of such a technique would be processing all the keywords searched for in Google's search engine\footnote{https://www.google.com}, and then grouping the results by keyword to count how many times a keyword was used in searches (for trend analysis purposes). Between batch and stream processing, batch processing is the older paradigm. Historically, Apache Hadoop \cite{borthakur2007hadoop, Hadoop, ApacheHadoop} was one of the first widely known frameworks for large-scale batch processing, created in 2005. Hadoop uses a programming model called MapReduce \cite{MapReduce}. This paradigm applies a map function that processes key/value pairs and generates sets of intermediate ones, ultimately merging all intermediate values
using a reducer function. To do this, the distributed computation framework reads and writes from disk on intermediate steps, which makes it inefficient for applications that reuse working sets of data across multiple parallel operations (\textit{e.g.}, iterative Machine Learning algorithms). Nearly a decade later, Apache Spark \cite{ApacheSpark, Spark} was released as the industry accepted successor.
Spark retains the scalability and fault tolerance of MapReduce but introduces Resilient Distributed Datasets (RDDs) \cite{SparkRDDs}. An RDD is a read-only collection of objects partitioned across a set of machines that can be rebuilt if a partition is lost.
Additionally, Spark processes data in random access memory (RAM), while Hadoop MapReduce persists data to the disk after a map or reduce action. Thus, Spark achieves lower processing times.

@@

In order to provide a streaming solution, Apache Spark created the Spark Streaming \cite{SparkStreaming} module. Being a batch system in its core, Spark implements streaming as micro-batching \cite{SparkStreamingPaper}. This approach handles streaming data by grouping incoming events into very small batches. Then, to process the entire defined window of data, it joins all micro-batches into a larger one and performs the desired computations. However, this creates an artificial barrier that does not truly exist in the streaming context. More than conceptually different from a true stream processing system, processing a stream in a batch fashion, even if in small ones, presents some issues. First, since we are indeed working with batches, results will never be updated in real-time per event being processed. For example, if we use a batch size of 100 events, the system would only produce accurate results every 100 events. Usually, to handle system inactivity, these micro-batch systems also have the notion of updating the aggregations with incomplete batches, if no event has entered the system after a pre-specified time period has elapsed. As such, micro-batching was created as a natural implementation of stream processing using the available batch processing systems of the time, but can not be considered true stream processing.

The Venn diagram in Figure \ref{fig:stream-superset} \cite{batch-is-a-special-case-of-streaming} illustrates the relation of both processing techniques. A data stream is unbounded or infinite whereas a bounded dataset has a beginning and an end. Using these definitions, the concept of an unbounded data stream contains that of a bounded dataset. That is, from an infinite stream, it is possible to fix left and right limits to obtain a bounded and finite dataset. This relation is visible in Figure \ref{fig:unbounded-bounded-superset} from the same blog post.

@@

In the next section, we will cover exactly how an unbounded dataset (also known as a data stream) can be converted into a bounded dataset with the use of windows.

@@

In Section \ref{sec:stream-superset}, we have defined a data stream as an unbounded dataset. But what exactly are the issues of working with an infinite stream of data? Why might we need to convert it into a bounded dataset and how to do so? In this section, we will analyze such issues and present sliding windows as the solution.

@@

Botan \emph{et al.} \cite{Botan-SECRET} define a window as:
\begin{definition}
A window over a stream $\mathbb{S}$ is a finite subset of $\mathbb{S}$
\end{definition}
Windows can be categorized by two dimensions: the window size $W$ and the sliding step size $S$. Windows can be defined as tuple or time-based. Tuple-based windows are windows whose size is defined by \emph{number} of items contained in the window, whereas time-based windows are windows whose size is defined by the total \emph{time} a window can store. For instance, a tuple-based window of size $10^3$ tuples will store exactly $10^3$ items. In contrast, a time-based window storing the last 5 minutes of events can contain a varying number of tuples but will always keep the events arrived in the last 5 minutes. A window slides over incoming data and groups it, enabling us to compute aggregations over a defined and bounded set of events. The step size \textit{S} of a windowing technique is defined as the distance between consecutive windows \cite{Botan-SECRET}. This means that two consecutive windows W\textsubscript{1} and W\textsubscript{2} are exactly \textit{S} units away. As mentioned, these units can be units of time or a number of tuples.

@@

% tumbling: step = window size
A tumbling window is a window with a stepping size as large as the window, \textit{i.e.}, $S = W$. Hence, tumbling windows partition the stream into non-overlapping chunks with the same size. Since the intersection of different tumbling windows is always empty, each event is said to belong to exactly one tumbling window. An example of a tumbling window can be seen in Figure \ref{fig:tumbling-window} with a tuple-based window of size \textit{W} of three and an equally large step size \textit{S}.
\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=1.1]{figures/tumbling.pdf}
      \caption[Tumbling window]{Tumbling window, W=S=3}
      \label{fig:tumbling-window}
    \end{center}
\end{figure}

@@

An aggregation --- also known as a fold, reduce or accumulate function --- is an operation that reduces a collection of values into a single one. Examples of such are the computation of dataset sums, minima, averages or the number of distinct elements (cardinality). An aggregation may be exact or approximate. Exact aggregations compute a value from a sequence of elements, where no error or loss of resolution exists. On the other hand, approximate aggregations have an error margin to the associated aggregation value which is usually controllable by adjusting a set of parameters. 
Approximate aggregations are most useful in situations where memory is limited, the volume of data to process is large and/or computing the exact aggregation can not be done in useful time. Thus, approximate aggregators trade precision of results for efficient time and space usage. In Section \ref{sec:pds}, we explore approximate aggregators that make such a trade-off.

Aggregations are present in most programming languages and frameworks. Quoting from the Java programming language documentation on the \textit{Stream.reduce()} method \cite{Java13StreamReduce}: 
\textit{"A reduction operation (also called a fold) takes a sequence of input elements and combines them into a single summary result by repeated application of a combining operation"}. Similarly, the MapReduce \cite{MapReduce} paradigm consists of first applying \textit{map} steps to the collection of data and then a \textit{reduce} one. \textit{Map} tasks will transform each item as desired and the \textit{reduce} task will merge the collection into a single value. Reduce methods make use of a \textit{combine} function and recursively apply it. The \textit{combine} function takes the current aggregation value and combines it with the next item in the set, returning the new aggregated state. Doing this recursively for each item in the collection results in a single value: the aggregate value. The \textit{combine} function is from here onwards represented as $\oplus$ and the identity or neutral element of the aggregation as $\overline{\theta}$.

@@

While some aggregations can be computed on an infinite dataset, often we want to compute these over a finite one. As we have seen in Section \ref{sec:windows}, we can obtain a bounded dataset from a data stream by applying windowing techniques. When doing so, we obtain a finite dataset denoted as a sliding window and compute aggregations over it. Tangwongsan \emph{et al.} define such aggregations over sliding windows as Sliding Window Aggregations (SWAGs) \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}. As an example of a SWAG, consider the data center temperature monitoring problem from Section \ref{sec:windows}. In this example, we wanted to compute the average temperature per rack in the data center in the last 5 minutes. If we translate this SWAG into streaming SQL following a similar syntax to the one provided by Apache Flink \cite{ApacheFlink} we would get:
\begin{verbatim}
SELECT rack_id, AVG(temperature)
FROM Temperatures
GROUP BY TUMBLE(timestamp, Time.minutes(5), rack_id)
\end{verbatim}
In the query above, the
\texttt{TUMBLE(timestamp, Time.minutes(5), rack\_id)} statement groups incoming data using time-based tumbling windows of 5 minutes duration, using the \texttt{timestamp} field to evict events not in this time interval. The resulting list of rows has \texttt{rack\_id} as its key and as corresponding value a list of temperatures, from the past 5 minutes. Finally, for each rack and its list of temperatures, we compute an \texttt{AVG} aggregation, resulting in the average temperature per rack in the past 5 minutes. Note that this is very similar to any aggregation made in a standard Relational Database Management System (RDBMS). The main difference is that in standard SQL the \texttt{GROUP BY} clause includes only the columns on which the data will be grouped while in streaming SQL the \texttt{GROUP BY} also specifies the type of window used (\textit{e.g.}, in this case, a tumbling window).

@@

Two of the most basic SWAG algorithms are Recalculate-From-Scratch (RFS) and Subtract-On-Evict (SOE). Recalculate-From-Scratch (RFS) is the simplest and most general SWAG algorithm. As the name suggests, this algorithm recomputes the aggregation over the entire window for each update. For instance, the computation of a sum over a sliding window using RFS means that whenever the window changes all elements are summed. This approach works for every aggregation. However, given \textit{n} as the window size, this algorithm takes \textit{O(n)} space and time complexity which makes it unfit in most scenarios of large scale data processing. Recall that the \textit{combine} function is represented by $\oplus$ and the identity or neutral element of the aggregation by $\overline{\theta}$. Pseudocode for the Recalculate-From-Scratch algorithm is shown in Algorithm \ref{pseudo:rfs}, assuming all functions have access to the \textit{vals} queue which holds the window contents.

@@

Subtract-On-Evict (SOE) on the other hand, is a SWAG algorithm that incrementally computes the aggregation value. For each element inserted or evicted from the window, the aggregation value is updated without recomputing the entire aggregation like in RFS. SOE relies on expiring the evicted elements from the aggregation state. Hence, SOE only works for \textit{invertible} aggregations, that is, a function $\ominus$ exists such that $(x \oplus y) \ominus y = x$, for all $x$ and $y$. To illustrate the differences between SOE and RFS, consider the computation of a sum aggregation over a sliding window. The total sum of the window will be the aggregation state or value. When inserting an element, the updated total window sum equals the current total sum plus ($\oplus$) the new element. The sum aggregation is considered \textit{invertible} because, when evicting an element, the new aggregation state is the total sum computed thus far minus ($\ominus$) the evicted element. With window size of \textit{n}, SOE has \textit{O(n)} space complexity and \textit{O(1)} time complexity. However, aggregations are not always invertible thus making this algorithm very restrictive and not always applicable. Algorithm \ref{pseudo:soe} shows pseudocode for SOE assuming all functions have access to the window contents through the \textit{vals} queue and access to the current aggregation state \textit{agg}.

@@

As it turns out, besides \textit{invertibility}, other properties allow us to group aggregations in different families. Tangwongsan \emph{et al.} construct Table \ref{tbl:aggregations-properties} to summarize the aggregation families and satisfying properties \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}. To understand these properties, we must be familiar with the \textit{combine} and \textit{query} functions. The already mentioned \textit{combine} function merges the incoming event with the incrementally computed aggregation state. The \textit{query} function is responsible for returning a result for a given query according to the current aggregation state. For instance, consider once again the use case of computing a sum aggregation over a window of data. The \textit{combine} function would be the arithmetic operator \textit{+} and would merge the current aggregation state --- \textit{i.e.}, the current sum value --- with the new incoming event. The \textit{query} function would return the current aggregation state --- \textit{i.e.}, return the total sum computed. Based on the \textit{combine} and \textit{query} functions, the following are formal definitions of these properties. An aggregation is said to have:

\begin{enumerate}
    \item  an \textbf{\textit{invertible combine}}, if a function $\ominus$ exists such that $(x \oplus y) \ominus y = x$, for all $x$ and $y$;

    \item  an \textbf{\textit{associative combine}}, if $(x \oplus y) \oplus z = x \oplus (y \oplus z)$, for all $x$ and $y$;

    \item  a \textbf{\textit{commutative combine}}, if $x \oplus y = y \oplus x$, for all $x$ and $y$;

    \item  a \textbf{\textit{size-preserving combine}}, if the aggregation state resulting from $\oplus$ always occupies the same space in memory;

    \item  a \textbf{\textit{unary query}}, if the query does not need any parameters --- \textit{e.g.}, a membership test with a Bloom Filter (detailed in Section \ref{sec:bloom}) would require the element to test membership for as a parameter so it is not unary; returning the total sum is unary since it does not require any parameters.
\end{enumerate}

@@

    \end{center}
    \caption[Aggregation properties]{Aggregation properties. Checkmarks (\checkmark), crosses ($\times$), and question marks (?) indicate a property is true for all, false for all or false for only some aggregations of that family, respectively \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}}
  \label{tbl:aggregations-properties}
\end{table}

@@

In Section \ref{sec:windows}, we discussed windowing mechanisms that allowed us to convert unbounded data streams (or infinite datasets) to finite datasets.

In Section \ref{sec:aggregations}, we introduced aggregations over datasets. We gave examples of computable aggregations over unbounded datasets at first, but then discussed sliding window aggregations, which make use of windowing techniques to obtain a finite dataset, on which the aggregation is computed. We further discussed algorithms used for the efficient computation of sliding window aggregations. Lastly, we presented a set of aggregation properties.

### 2020-07-06 chapters/introduction.tex
Final changes and submission

\section{Context} \label{sec:context}
In the past decade, applications have become increasingly data-driven, placing data at the center of application design. For different use cases, the data is processed in different ways for different purposes. For instance, e-commerce platforms such as Alibaba\footnote{https://alibaba.com} need to process a large number of daily transactions while ensuring that sales run smoothly and that products are delivered to customers' homes. Streaming services, like YouTube\footnote{https://youtube.com} and Netflix\footnote{https://netflix.com}, try to guarantee that media content reaches up to millions of users simultaneously. Social networks are responsible for generating large volumes of data of which,  Twitter\footnote{https://twitter.com} for example, generates more than 500 million \textit{tweets} (posts in Twitter) per day\footnote{https://blog.twitter.com/engineering/en\_us/a/2013/new-tweets-per-second-record-and-how.html}. \textit{Tweets} may contain text, media content or both. Cybersecurity applications are yet another example of time-sensitive, data-driven applications. The entities that work in these use cases need to monitor user accesses and their actions in the network, where timely detection of intruders is critical to prevent them from tampering with the underlying system.

@@

In either processing scenario, we usually want to apply operations on data, for example, to compute the maximum value, an average or a count of distinct elements. Operations that produce a single result when applied to datasets are commonly known as aggregation operations. In stream processing, we apply these aggregations to data streams to obtain valuable information. However, since a data stream is, effectively speaking, an unbounded dataset, if our aggregation requires a finite domain, there is the need to apply windowing techniques. A thorough analysis of stream windowing techniques is done in Chapter \ref{chap:background} of this document but for now it suffices to say that windowing is a mechanism to select a fine portion of an unbounded data stream, typically by defining a time or tuple-based window size --- \textit{e.g.}, events from the last three days or the last million events, respectively.

@@

\section{Motivation} \label{sec:motivation}
For any organization, the quality of their provided services is paramount. Therefore, it is common for companies to have auxiliary systems monitoring their services, testing their usage rate, availability and performance, just to name a few. The motivation behind this thesis is to produce a system capable of real-time monitoring a stream of data so that changes in the underlying stream are caught early, preventing a fall in the performance of the system analyzing said data stream.

@@

    \item explainable alerts
\end{itemize}
The hypothesis will be tested and evaluated on the previous criteria. Experiments will be made using multiple synthetic datasets and a real dataset from the financial fraud space.

@@

In Chapter \ref{chap:statement}, we define the problem this thesis aims to solve, the issues with state of the art solutions, our proposed solution, the assumptions made on it and the research questions we intend to answer.

In Chapter \ref{chap:my-work}, we discuss initial approaches and reasons to discard them before presenting our proposed solution.

>>>>

\section{Context} \label{sec:context}
In the past decade, applications have become increasingly data-driven, placing data at the center of application design. For different use cases, the data is processed in different ways for different purposes. For instance, e-commerce platforms such as Alibaba\footnote{https://alibaba.com} need to process a large number of daily transactions while ensuring that sales run smoothly and that products are delivered to customers' homes. Streaming services, like YouTube\footnote{https://youtube.com} and Netflix\footnote{https://netflix.com}, try to guarantee that media content reaches up to millions of users simultaneously. Social networks are responsible for generating large volumes of data. Twitter\footnote{https://twitter.com}, for example, generates more than 500 million \textit{tweets} (posts in Twitter) per day\footnote{https://blog.twitter.com/engineering/en\_us/a/2013/new-tweets-per-second-record-and-how.html}. \textit{Tweets} may contain text, media content or both. Cybersecurity applications are yet another example of time-sensitive, data-driven applications. The entities that work in these use cases need to monitor user accesses and their actions in the network, where timely detection of intruders is critical to prevent them from tampering with the underlying system.

@@

In either processing scenario, we usually want to apply operations on data, for example, to compute the maximum value, an average or a count of distinct elements. Operations that produce a single result when applied to datasets are commonly known as aggregations. In stream processing, we apply these aggregations to data streams to obtain valuable information. However, since a data stream is, effectively speaking, an unbounded dataset, if our aggregation requires a finite domain, there is the need to apply windowing techniques. A thorough analysis of stream windowing techniques is done in Chapter \ref{chap:background} of this document but for now it suffices to say that windowing is a mechanism to select a fine portion of an unbounded data stream, typically by defining a time or tuple-based window size, \textit{e.g.}, events from the last three days or the last million events, respectively.

@@

\section{Motivation} \label{sec:motivation}
For any organization, the quality of their provided services is paramount. Therefore, it is common for companies to have auxiliary systems monitoring their services, testing their usage rate, availability and performance, just to name a few. The motivation behind this thesis is to produce a system capable of real-time monitoring a stream of data so that changes in the underlying stream are caught early, preventing a fall in the performance of another system analyzing said data stream.

@@

    \item explainable alerts
\end{itemize}
The hypothesis will be tested and evaluated on the previous criteria. Experiments will be made using multiple synthetic datasets and real data from the financial fraud space, provided by Feedzai.

@@

In Chapter \ref{chap:statement}, we define the problem this thesis aims to solve, the issues with state of the art solutions, our proposed solution, the assumptions made with it and the research questions we intend to answer.

In Chapter \ref{chap:my-work}, we discuss initial approaches and the reasons that led us to discard them, before presenting our proposed solution.
